### 2016-01-28 paper/chemokine_main_revised.tex
In process of drafting Response 3

\documentclass[preprint,10pt,authoryear]{elsarticle}

@@

% cite package, to clean up citations in the main text. Do not remove.
%\usepackage{cite}
\usepackage{natbib}

@@

@@

\bibliographystyle{elsarticle-num-names}

@@

\newcommand{\removed}[1]{{\color{dkred}\sout{#1}}}
\newcommand{\drew}[1]{{\color{dkgreen}#1}}
\newcommand{\fred}[1]{{\color{dkblue}#1}}
\newcommand{\steph}[1]{{\color{dkpurple}#1}}

@@

The model is parameterized with values taken from the literature when available.  Because chemokine secretion rates are central to our model and appropriate data values are not available, we estimate the parameters using data from physical experiments and an earlier ODE model \citep{Mitchell2011}.   To obtain the physical data, wells of human epithelial cells were infected with one of the three strains and then measured for virus and cytokine concentrations over a 48-hour period.   Of the cytokines tested, only the chemokines CXCL10 (IP-10) and CCL5 (RANTES) were expressed differently across the three influenza strains.   IP-10 and RANTES are both chemokines that normally attract T cells to the FOI \citep{Hoji2005, Groom2011a}.  These results suggest that the two chemokines may contribute to the immune system's varying ability to control different strains of influenza, and were thus included in our model.

The spatial model and the supporting experiments reveal several interesting and underappreciated phenomena.  The model identified physical constraints on T cells' ability to clear the rapidly replicating pH1N1 infection. Because T cells in the model use the chemokine gradient to locate the FOI, they tend to cluster in areas of high chemokine concentration. The model suggests that the chemokine response of the host cells can hinder T cells’ ability to find and clear infected cells in the presence of the rapidly replicating pandemic strain. A sensitivity analysis of the model parameters isolated the effects of each parameter in the model.  The analysis shows that most model parameters can vary across a broad range of values without strong effect on predicted outcomes.  Further analysis shows model predictions remain unchanged within broad categories of parameter type (e.g., all parameters related to chemokines, parameters related to viral kinetics, etc.).  For example, the chemokine-related parameters are stable in that they do not affect the model behavior within biologically plausible ranges. In contrast, parameters related to viral kinetics significantly alter model predictions.  Finally we examine the `window of control' which establishes an upper bound on the time it takes T cells to arrive at the FOI while still clearing the infection.  We find that this value is consistent with our sensitivity analysis and helps explain the challenge of controlling the highly virulent pH1N1 strain.

@@

The model is initialized with a single infected cell.  After the incubation period, the infected cell begins secreting virus and chemokine at rates determined by the virus strain and chemokine type (Table \ref{tab:strains}).  Virus from expressing cells diffuses locally, infecting neighboring cells. Chemokine diffuses from secreting cells, creating a spreading region of stimulation around the FOI. After five simulated days, representing lymph node stimulation and T cell proliferation, activated T cells exit the lymph node at a constant rate and travel through the vasculature to the tissue as described above.  These cell and molecular interactions and state transitions are depicted in Figure~\ref{fig:systemchart}.

@@

In the model, epithelial cells are stationary and are described by one of five sequential states: \texttt{healthy}, \texttt{virus-incubating}, \texttt{virus-expressing}, \texttt{apoptotic}, and \texttt{dead} \citep{bachem1996simulated, Beauchemin2005, Mitchell2011}. \texttt{Healthy} cells remain unchanged unless infected by virus. Once infected, the cell transitions from \texttt{incubating} to \texttt{expressing} after a 10 hour incubation delay. \texttt{Expressing} cells secrete virus and chemokine for a fixed 16.7 hours and then die. \texttt{Expressing} cells initiate apoptosis sooner if they are contacted by activated T cells. \texttt{Apoptoic} cells continue to secrete virus and chemokine and then die after one hour. \texttt{Dead} cells take up space and do not regenerate over the course of an infection.  Cell regrowth is not implemented as the rate of regrowth is unknown.

T cells are described by two states: \texttt{circulating} and \texttt{chemotaxing}.    T cells emerge from the lymph node at five days post infection (p.i.) at a rate of 1,257 cells per hour. T cell travel time from the lymph node to a random location on the lung's surface is six seconds.  If chemokine is not encountered, the circulating T cell returns to a new location in the lung after another six seconds.  If a circulating cell encounters chemokine, it changes state to \texttt{chemotaxing} and follows the chemotactic gradient to the FOI. Circulating T cells decay exponentially with an average lifespan of three days.  Chemotaxing T cells follow the gradient through the two-dimensional lung endothelium, inducing apoptosis when they encounter expressing epithelial cells. Chemotaxing T cells decay exponentially with an average lifespan of two hours.

In addition to T cells, the model contains two types of particles: virus and chemokine. Both are produced at constant rates by expressing epithelial cells.  Virus diffuses through the lung tissue, infecting healthy cells at a rate proportional to the virus concentration at the location of the cell. Chemokine diffuses across the tissue but has no direct effect beyond recruiting T cells. Both particle types decay exponentially.  IgM is modeled by increasing the viral decay rate by a factor of ten after the fourth day.

Because we are interested in the interplay between the virus and the induced immune response, we infected human epithelial cells with the three different strains of influenza \textit{in vitro} and measured the resulting cytokine and chemokine responses.  We extended a previously published differential equation model \citep{Mitchell2011} to obtain values for per-cell chemokine production rates.  T cell production rates were derived from measured replication rates \textit{in vitro} \citep{Miao2010a} using another differential equation model. T cell production is assumed to be constant after day 5 \citep{MartIn-Fontecha2003}.

@@

We estimate chemokine production rates, $r$, by adapting the delay differential equation model of influenza infection described in \citep{Mitchell2011} Eq. 1 by adding one new equation ($\dot{C}=r I_{1 \tau_3}-dC$) to model chemokine production.  Initial population sizes and parameter values are taken from the previous study.
{\footnotesize
\begin{equation*}
\begin{aligned}
\dot{T} &= - \beta T V \\
\dot{I_1} &= \beta T V - \beta T_{\tau_1}V_{\tau_1} \\
\dot{I_2} &= \beta T_{\tau_1}V_{\tau_1} - \delta I_2 \\
\dot{V} &= \frac{p}{1+eF} I_2  - \beta T V  \\
\dot{F} &=  I_{1 \tau_2} \\
\dot{C} &= r I_{1 \tau_3} - d C \\
\end{aligned}
\tag{Eq. 1}
\label{eq:dde}
\end{equation*}
\vspace{.05in}
}

$\tau$ subscript variables denote delay terms, signifying the value is the population quantity in existence at time $t - \tau$.  Table S2 summarizes population and parameter values and descriptions.  Strain-specific values for $r$ were found by fitting the equations to the experimental data in Table S1.  The results of these fits are discussed in the main paper and shown in Table 1.  IP-10 and RANTES secretion rates are aggregated in the spatial model (S2.2).

@@

$N_{c}$ is the number of circulating activated antigen-specific CD8 T cells, $N'_{f}$ is the number of circulating T cells that have found and exit into a region of lung tissue expressing chemokines, $N_{f}$ is the number of circulating T cells that have found an infected region. $T$ is the number of uninfected target cells, $I$ is the number of productively infected cells, and $V$ is the viral titer in serum. The infected region is assumed to be of radius $r$ and is within a region expressing chemokines of radius $r'$ ($r  < r'$). The lung is modeled as a circular region with radius $R$. The area of the infected region is equal to the area of an infected cell (of radius $r_{cell}$) multiplied by the number of infected cells. See Fig. 1 for more details about the model.  The area of the region expressing chemokine was found to be related non-linearly to the number of infected cells ($\pi r'^{2} = a \sqrt{I} + b$) where $a$ and $b$ are constants that depend on the viral strain.  $a$ and $b$ were fit using 12 experimental runs of a spatial model implemented in CyCells.

@@

The average time an exiting CD8 T cell takes to migrate to the infected region is equal to the difference in the radii between the two regions ($r' - r$) times the T cell speed, $v_{tcell}$.  Circulating cells that are in the chemokine expressing region ($N'_{f}$) move into the infected region after performing chemotaxis for that time. Target cells ($T$) become infected by virus at rate $\beta TV$, where $\beta$ is the rate constant characterizing infection. Infected cells ($I$) die at rate $\delta$ in addition to being lysed by T cells ($N_{f}$) at a rate $k_{e}$. Finally the viral titers ($V$) increase due to production of virus at rate $p$ by infected cells. Virus is also cleared due to uptake by infected cells (at a rate $- \beta TV$) and due to antibody (at a rate $\gamma (t)$ that changes after 5 days post infection). The initial viral titer was initialized to 10,000 PFUs and the initial number of target cells was one million. The initial number of infected cells is assumed to be zero. Parameter values are listed in Table S3.  This ODE was fit to data taken from \citep{Miao2010a} using Matlab's \texttt{nlinfit} function in order to obtain a value for $\sigma$.  The final value was found to be 1,257 per hour.

@@

In an effort to obtain per-cell chemokine and cytokine production rates, human epithelial cells were infected with influenza \textit{in vitro}.  Epithelial cell culture and supernatant collection was performed as described in \citep{Mitchell2011}.  Briefly, undifferentiated human tracheal epithelial cells (University of Miami) were cultured for 4 weeks to achieve fully differentiated confluent monolayers on collagen-coated transwell inserts, or commercial differentiated human bronchial epithelial cells (EpiAirway Tissue, MatTek Corp., Ashland, MA) used immediately upon receipt, were infected at an MOI of 0.01 with either seasonal H1N1 virus A/New Caledonia/20/99 (sH1N1), the 2009 H1N1 pandemic strain A/California/04/09 (pH1N1), or avian H5N1 virus A/Hong Kong/483/97 (aH5N1) derived from a fatal human infection.  Basal media was collected from previously undisturbed triplicate or quadruplicate wells at 0, 6, 10, 12, 16, 20, 24, 30, 36, 42, 48, and 72 hours after infection, and stored at -80C until assay.  Subsequently, apical fluid for virus secretion was collected before and after treatment of the monolayer with protease (Pronase, Sigma) to optimize the collection of infectious virus \citep{Mitchell2011}.  Quantitative viral culture was performed by standard plaque assay.  Quantitative chemokine levels were performed in 30 $\mu$L aliquots for a panel of chemokines (IL-8, MCP-1, MIP-1$\alpha$, MIP-1$\beta$, RANTES, IP-10, eotaxin) and cytokines (interferon-gamma, IL-1$\alpha$, IL-1$\beta$, IL-1RA, IL-2, IL-3, IL-4, IL-6, IL-10, IL-12p40, IL-15, IL-17, TNF$\alpha$) (Luminex Assay®, Luminex Corp.) and reported as ng/mL basal media sampled from a total volume of 4 mL.  Only IP-10, RANTES, and TNF$\alpha$ showed increases in production (Table S1).  TNF activity is not incorporated in the model.  Data for other chemokines and cytokines is not shown.

@@

Because the CyCells model is stochastic, we conducted 50 runs of the model for each of the three influenza strains, each run initialized with a unique random seed (Fig.~\ref{fig:variance}).  For the avian and seasonal strains, all model runs cleared the virus by day 10.  Conversely, each simulation of the 2009 pandemic influenza led to uncontrolled infection.

@@

Plaque growth in the model is illustrated in Figure \ref{fig:cycells} and Videos S1-S3.  Because the virus particles are an order of magnitude larger than the chemokine molecules, they diffuse more slowly, according to the Stokes-Einstein equation.  However, chemokine molecules decay more quickly than the virus (Table~\ref{tab:parameters}).  In the model these countervailing characteristics produce similar spatio-temporal profiles for the two particle types (Fig.~\ref{fig:cycells}). Before day 4 the plaque is populated primarily by virus-incubating and virus-secreting cells, with a small percentage of dead cells (Panel A). Over time, cells in the plaque's interior die, and active cells comprise a decreasing proportion of the plaque. T cells arrive at day 5 and begin killing the virus-secreting cells (Panel B). By day 6 many expressing cells have been eliminated and the plaque is populated by dead cells (Panel C).

@@

Finally, the spatial nature of our model reveals that chemokines can diffuse more slowly than the rate infected cells and virus expand, thus misdirecting the T cells.  It takes time for infected cells to begin producing chemokine, while the preexisting areas of high chemokine density are slow to decay.  Thus, T cells whose movements respond to the spatial layout of the chemokine gradient can become trapped, failing to locate new regions of infected cells in the growing plaque.  This effect is most pronounced in the pH1N1 strain, where the plaque expands more quickly than the chemokine can diffuse.

% Taken together, the delayed response and low proportion of virus secreting cells contribute strongly to the failure of T cells clearing the infection.  Furthermore, increasing chemokine production fails to guide T cells to the actively secreting epithelial cells in the runaway pH1N1 infection (Fig. S5), suggesting that the effect of high viral production rates dominate increased chemokine production rates.

@@

We conducted a sensitivity analysis of sixteen model parameters to identify those that have an effect on determining model outcomes (S2.3). We studied all but five of the parameters listed in Tables~\ref{tab:strains} and \ref{tab:parameters}, excluding T cell radius, T cell sensitivity to chemokine, T cell onset time, IgM onset time, and the viral production rate.  T cell radius is known \citep{abbas2011cellular} and does not factor strongly into the spatial model's implementation.  T cell sensitivity was set to mimic a realistic threshold of response to cytokine (see S2.1 for a more complete explanation of how T cell sensitivity was determined).  The two onset parameters are known \citep{Diamond2003}.  Viral production is an independent variable, as discussed above, and determined by each of the three strains \citep{Mitchell2011}.

Each of the remaining sixteen parameters was varied independently (over multiple orders of magnitude when appropriate) for each viral strain, and the model was rerun once for each parameter setting.  Results were compared for the quantitative size of the peak infection and the number of uncleared infected cells at day 10 p.i.  Based on these runs, we classified each parameter for each strain as follows (Table~\ref{tab:sensitivity}): \textit{stable} parameters are those that do not affect the model behavior over any of the different values tested; \textit{bounded stable} parameters are stable throughout a reasonable range of values but model behavior diverges at some threshold; \textit{peak change} parameters affect the peak size of the infection, but do not change the ultimate number of infected cells at the end of the run; and \textit{sensitive} parameters alter the outcome of the simulation for any variation within a biologically plausible range.

Table \ref{tab:sensitivity} shows that four parameters are sensitive: viral response to IgM, infectivity, viral decay rate, and viral diffusion.  In addition, Figure \ref{fig:variance} shows that viral production rate is a sensitive parameter due to the difference between the three panels.  The four parameters (infectivity, viral decay rate, viral diffusion rate, and viral production rate) are specific viral characteristics.  Further, viral response to IgM, as coded in the model, directly affects viral decay rate and can be included in this group.  None of the T cell parameters nor the chemokine parameters have a strong effect on the model's prediction of control or lack of control of viral replication.  This result supports the hypothesis that virus dynamics dominate the course of an influenza infection.

@@

Our sensitivity analysis suggests that viral parameters are the most sensitive and have the largest effect on the the course of the infection (Table~\ref{tab:sensitivity}).  Here we examine the role of T cells by assessing multiple factors that control the overall T cell response.  We seek to identify the maximum amount of time virus-secreting cells may produce virus while still allowing the infection to be cleared.  We name this threshold the `window of control'.  This window can be defined as the combination of the the time of T cell arrival at a virus-secreting cell ($T_{arr}$), the time it takes for the T cell to induce apoptosis ($T_{kill}$) and the time for the infected cell to apoptose ($T_{apop}$).  We examine how much virus a virus-secreting cell can produce inside this window.  Values above one suggest a growing infection, while values below one suggest infection control.

@@

$p$, $T_{kill}$, and $T_{apop}$ are model parameters (Table \ref{tab:parameters}).  $E$ incorporates both spatial effects of viral diffusion and temporal effects of viral decay.  As the infected plaque expands, some secreted virus diffuses to cells that are already infected or dead (the existing plaque) and do not contribute to new infections.  This effect is minimal in the beginning when the edge of the plaque is small and tightly curved.  As the plaque expands, its radius grows and the edge approaches the limiting case of a straight line when 50\% of the virus could be expected to diffuse over healthy cells and 50\% over infected or dead cells.  In the presence of healthy cells, virions infect the cells at a rate of $0.5/h$ and decay at a rate of $0.42/h$ in the presence of IgM after day 4 p.i.  (Table \ref{tab:parameters}).  This suggests that 50\% of free virus contacts healthy cells and 54\% of that virus succeeds in infecting a healthy cell, implying that $E \approx 0.27$ (see S2.4 for a more complete explanation of how $E$ was determined).  Solving for $T_{arr}$ where $R_0 < 1$ gives:

>>>>

\documentclass[preprint,10pt,numbers]{elsarticle}

@@

% cite package, to clean up citations in the main text. Do not remove.
%\usepackage{cite}
\usepackage[]{natbib}

@@

\usepackage{xr}
\externaldocument{chemokine_supplement_revised}

@@

%\bibliographystyle{elsarticle-num}
\bibliographystyle{elsarticle-num-names}
%\bibliographystyle{plos2009}

@@

\newcommand{\removed}[1]{{\color{dkred}\sout{#1}}}
\newcommand{\new}[1]{{\color{dkgreen}#1}}
%\newcommand{\removed}[1]{}
%\newcommand{\new}[1]{#1}
\newcommand{\drew}[1]{{\color{dkgreen}#1}}
\newcommand{\fred}[1]{{\color{dkblue}#1}}
\newcommand{\steph}[1]{{\color{dkpurple}#1}}

@@

The model is parameterized with values taken from the literature when available.  Because chemokine secretion rates are central to our model and appropriate data values are not available, we estimate the parameters using data from physical experiments and our previously published ODE model \citep{Mitchell2011}.

To obtain the physical data, wells of human epithelial cells were infected with one of the three strains and then measured for virus and cytokine concentrations over a 48-hour period.   Of the cytokines tested, only the chemokines CXCL10 (IP-10) and CCL5 (RANTES) were expressed differently across the three influenza strains.   IP-10 and RANTES are both chemokines that normally attract T cells to the FOI \citep{Hoji2005, Groom2011a}.  These results suggest that the two chemokines may contribute to the immune system's varying ability to control different strains of influenza, and were thus included in our model.

The analyses of the spatial model reveal several interesting and underappreciated phenomena.  The model identified physical constraints on T cells' ability to clear the rapidly replicating pH1N1 infection. Because T cells in the model use the chemokine gradient to locate the FOI, they tend to cluster in areas of high chemokine concentration. The model suggests that the chemokine response of the host cells can hinder T cells’ ability to find and clear infected cells in the presence of the rapidly replicating pandemic strain. A sensitivity analysis of the model parameters isolated the effects of each parameter in the model.  The analysis shows that most model parameters can vary across a broad range of values without strong effect on predicted outcomes.  Further analysis shows model predictions remain unchanged within broad categories of parameter type (e.g., all parameters related to chemokines, parameters related to viral kinetics, etc.).  For example, the chemokine-related parameters are stable in that they do not affect the model behavior within biologically plausible ranges. In contrast, parameters related to viral kinetics significantly alter model predictions.  Finally we examine the `window of control' which establishes an upper bound on the time it takes T cells to arrive at the FOI while still clearing the infection.  We find that this value is consistent with our sensitivity analysis and helps explain the challenge of controlling the highly virulent pH1N1 strain.

@@

The model is initialized with a single infected cell.  After the incubation period, the infected cell begins secreting virus and chemokine at rates determined by the virus strain and chemokine type (Table \ref{tab:strains}).  Virus from expressing cells diffuses locally, infecting neighboring cells. Chemokine diffuses from secreting cells, creating a spreading region of stimulation around the FOI. After four simulated days, Immunoglobulin M (IgM) is introduced into the model by increasing the viral decay rate.  After five simulated days, representing lymph node stimulation and T cell proliferation, activated T cells exit the lymph node at a constant rate and travel through the vasculature to the tissue as described above.  These cell and molecular interactions and state transitions are depicted in Figure~\ref{fig:systemchart}.

We include only parameters that directly address the role of T cells and T cell migration because our study focuses on the role of T cells and the chemokine effects of T cells in influenza virus infection. Thus, we included parameters that affected virus, T cells, and chemokines. With regards to the influenza, we chose parameters involving viral replication, infectivity in epithelial cells, and virus decay and diffusion rates based on our earlier study.  The chemokine decay rate, diffusion rate, and secretion rate parameters are relevant to dynamic chemokine gradients required for T cell chemotaxis.  Because we wanted to use our model to test the role of T cells and T cell migration in influenza clearance, we included multiple T cell parameters, such as T cell production rate, T cell death rate, T cell kill time, and T cell migration rate.  We added IgM because many studies have argued for the importance of antibody mediated virus clearance. However, because our study is not addressing the role of antibody in clearance of influenza, we did not test the full range of B cell responses directly by including them in our parameters.

Because we are interested in the interplay between the virus and the induced immune response, we infected human epithelial cells with the three different strains of influenza \textit{in vitro} and measured the resulting cytokine and chemokine responses.  We extended our previously published differential equation model \citep{Mitchell2011} to obtain values for per-cell chemokine production rates.  T cell production rates were derived from measured replication rates \textit{in vitro} \citep{Miao2010a} using another differential equation model. T cell production is assumed to be constant after day 5 \citep{MartIn-Fontecha2003}.

@@

In the model, epithelial cells are stationary and are described by one of five sequential states: \texttt{healthy}, \texttt{virus-incubating}, \texttt{virus-expressing}, \texttt{apoptotic}, and \texttt{dead} \citep{bachem1996simulated, Beauchemin2005, Mitchell2011}. \texttt{Healthy} cells remain unchanged unless infected by virus. Once infected, the cell transitions from \texttt{incubating} to \texttt{expressing} after a 10 hour incubation delay (Table \ref{tab:parameters}). \texttt{Expressing} cells secrete virus and chemokine for a fixed 16.7 hours and then die (Table \ref{tab:parameters}). \texttt{Expressing} cells initiate apoptosis sooner if they are contacted by activated T cells. \texttt{Apoptoic} cells continue to secrete virus and chemokine and then die after one hour (Table \ref{tab:parameters}). \texttt{Dead} cells take up space and do not regenerate over the course of an infection.  Cell regrowth is not implemented as the rate of regrowth is unknown.

T cells are described by two states: \texttt{circulating} and \texttt{chemotaxing}.    T cells emerge from the lymph node at five days post infection (p.i.) at a rate of 1,257 cells per hour (Models for Parameter Estimation, Table \ref{tab:parameters}). T cell travel time from the lymph node to a random location on the lung's surface is six seconds.  If chemokine is not encountered, the circulating T cell returns to a new location in the lung after another six seconds.  If a circulating cell encounters chemokine, it changes state to \texttt{chemotaxing} and follows the chemotactic gradient to the FOI. Circulating T cells decay exponentially with an average lifespan of four days.  Chemotaxing T cells follow the gradient through the two-dimensional lung endothelium, inducing apoptosis when they encounter expressing epithelial cells. Chemotaxing T cells decay exponentially with an average lifespan of two hours.  Justifications for these parameter values are listed in Table \ref{tab:parameters}.

In addition to T cells, the model contains two types of particles: virus and chemokine. Both are produced at constant rates by expressing epithelial cells.  Virus diffuses through the lung tissue, infecting healthy cells at a rate proportional to the virus concentration at the location of the cell. Chemokine diffuses across the tissue but has no direct effect beyond recruiting T cells. Both particle types decay exponentially.  IgM is modeled by increasing the viral decay rate by a factor of ten after the fourth day (Table \ref{tab:parameters}).

@@

To provide estimates of chemokine concentrations and secretion rates in lung tissue, chemokine levels were measured at 4-6h intervals during the first 48 h of infection in wells containing approximately one million human bronchial epithelial cells (Fig. \ref{fig:data}, Table \ref{tab:data}).  The dynamic viral loads at these intervals have been reported previously by us for cultures infected with seasonal H1N1 virus, pandemic H1N1 virus, and avian H5N1 virus \citep{Mitchell2011}.  IP-10 concentration increases were observed by 8h post-infection (p.i.), and RANTES by 16h p.i..

We estimate chemokine production rates, $r$, by adapting the delay differential equation model of influenza infection described in \citep{Mitchell2011} Eq. 1 by adding one new equation ($\dot{C}=r I_{1 \tau_3}-dC$) to model chemokine production.   Strain-specific values for $r$ were found by fitting the equations to the experimental data in Table~\ref{tab:data} using a genetic algorithm to minimize the log squared error between the model and the data while holding the rest of the parameter values constant \new{(Fig. S1, Table~\ref{tab:rmse})}.  Next, 1,000 bootstrapping runs using resampled residuals were performed on each modeled strain to generate confidence intervals over $r$.  The results of these fits are shown in Table 1 (viral secretion rates are from previous fits in \cite{Mitchell2011}).  IP-10 and RANTES secretion rates are aggregated in the spatial model (S2.2).

{\footnotesize
\begin{equation*}
\begin{aligned}
\dot{T} &= - \beta T V \\
\dot{I_1} &= \beta T V - \beta T_{\tau_1}V_{\tau_1} \\
\dot{I_2} &= \beta T_{\tau_1}V_{\tau_1} - \delta I_2 \\
\dot{V} &= \frac{p}{1+eF} I_2  - \beta T V  \\
\dot{F} &=  I_{1 \tau_2} \\
\dot{C} &= r I_{1 \tau_3} - d C \\
\end{aligned}
\tag{Eq. 1}
\label{eq:dde}
\end{equation*}
\vspace{.05in}
}
$\tau$ subscript variables denote delay terms, signifying the value is the population quantity in existence at time $t - \tau$.  Table \ref{tab:dde} summarizes population and parameter values and descriptions.

@@

$N_{c}$ is the number of circulating activated antigen-specific CD8 T cells, $N'_{f}$ is the number of circulating T cells that have found and exit into a region of lung tissue expressing chemokines, $N_{f}$ is the number of circulating T cells that have found an infected region. $T$ is the number of uninfected target cells, $I$ is the number of productively infected cells, and $V$ is the viral titer in serum. The infected region is assumed to be of radius $r$ and is within a region expressing chemokines of radius $r'$ ($r  < r'$). The lung is modeled as a circular region with radius $R$. The area of the infected region is equal to the area of an infected cell (of radius $r_{cell}$) multiplied by the number of infected cells. See Fig.\ref{fig:systemchart} for more details about the model.  The area of the region expressing chemokine was found to be related non-linearly to the number of infected cells ($\pi r'^{2} = a \sqrt{I} + b$) where $a$ and $b$ are constants that depend on the viral strain.  $a$ and $b$ were fit to 5 experimental runs of a spatial model implemented in CyCells ($R^2 = 0.997$, $P < 0.01$).

@@

The average time an exiting CD8 T cell takes to migrate to the infected region is equal to the difference in the radii between the two regions ($r' - r$) times the T cell speed, $v_{tcell}$.  Circulating cells that are in the chemokine expressing region ($N'_{f}$) move into the infected region after performing chemotaxis for that time. Target cells ($T$) become infected by virus at rate $\beta TV$, where $\beta$ is the rate constant characterizing infection. Infected cells ($I$) die at rate $\delta$ in addition to being lysed by T cells ($N_{f}$) at a rate $k_{e}$. Finally the viral titers ($V$) increase due to production of virus at rate $p$ by infected cells. Virus is also cleared due to uptake by infected cells (at a rate $- \beta TV$) and due to antibody (at a rate $\gamma (t)$ that changes after 5 days post infection). The initial viral titer was initialized to 10,000 PFUs and the initial number of target cells was one million. The initial number of infected cells is assumed to be zero. Parameter values are listed in Table \ref{tab:supplement}.  This ODE was fit to data taken from \citep{Miao2010a} using Matlab's \texttt{nlinfit} function in order to obtain a value for $\sigma$.  The final value was found to be 1,257 per hour.

@@

In an effort to obtain per-cell chemokine and cytokine production rates, human epithelial cells were infected with influenza \textit{in vitro}.  Epithelial cell culture and supernatant collection was performed as described in \citep{Mitchell2011}.  Briefly, undifferentiated human tracheal epithelial cells (University of Miami) were cultured for 4 weeks to achieve fully differentiated confluent monolayers on collagen-coated transwell inserts, or commercial differentiated human bronchial epithelial cells (EpiAirway Tissue, MatTek Corp., Ashland, MA) used immediately upon receipt, were infected at an MOI of 0.01 with either seasonal H1N1 virus A/New Caledonia/20/99 (sH1N1), the 2009 H1N1 pandemic strain A/California/04/09 (pH1N1), or avian H5N1 virus A/Hong Kong/483/97 (aH5N1) derived from a fatal human infection.  Basal media was collected from previously undisturbed triplicate or quadruplicate wells at 0, 6, 10, 12, 16, 20, 24, 30, 36, 42, 48, and 72 hours after infection, and stored at -80C until assay.  Subsequently, apical fluid for virus secretion was collected before and after treatment of the monolayer with protease (Pronase, Sigma) to optimize the collection of infectious virus \citep{Mitchell2011}.  Quantitative viral culture was performed by standard plaque assay.  Quantitative chemokine levels were performed in 30 $\mu$L aliquots for a panel of chemokines (IL-8, MCP-1, MIP-1$\alpha$, MIP-1$\beta$, RANTES, IP-10, eotaxin) and cytokines (interferon-gamma, IL-1$\alpha$, IL-1$\beta$, IL-1RA, IL-2, IL-3, IL-4, IL-6, IL-10, IL-12p40, IL-15, IL-17, TNF$\alpha$) (Luminex Assay®, Luminex Corp.) and reported as ng/mL basal media sampled from a total volume of 4 mL.  Only IP-10, RANTES, and TNF$\alpha$ showed increases in production (Table \ref{tab:data}).  TNF activity is not incorporated in the model.  Data for other chemokines and cytokines is not shown.

@@

Because the CyCells model is stochastic, we conducted 100 runs of the model with the same parameter set for each of the three influenza strains, each run initialized with a unique random seed (Fig.~\ref{fig:variance}).  We count the number of infected cells in the simulated plaque at every time point.  For the avian and seasonal strains, all model runs cleared the infection by day 10.  Conversely, each simulation of the 2009 pandemic influenza led to uncontrolled infection.

@@

Plaque growth in the model is illustrated in Figure \ref{fig:cycells} and Videos S1-S3.  Because the virus particles are an order of magnitude larger than the chemokine molecules, they diffuse more slowly according to the Stokes-Einstein equation.  However, chemokine molecules decay more quickly than the virus (Table~\ref{tab:parameters}).  In the model these countervailing characteristics produce similar spatio-temporal profiles for the two particle types (Fig.~\ref{fig:cycells}). Before day 4 the plaque is populated primarily by virus-incubating and virus-secreting cells, with a small percentage of dead cells (Panel A). Over time, cells in the plaque's interior die, and active cells comprise a decreasing proportion of the plaque. T cells arrive at day 5 and begin killing the virus-secreting cells (Panel B). By day 6 many expressing cells have been eliminated and the plaque is populated by dead cells (Panel C).

@@

Finally, the spatial nature of our model reveals that \new{the locations of peak chemokine concentration}\removed{chemokines} can \new{move}\removed{diffuse} more slowly than the rate infected cells and virus expand, thus misdirecting the T cells.  It takes time for infected cells to begin producing chemokine, while the preexisting areas of high chemokine density are slow to decay.  Thus, T cells whose movements respond to the spatial layout of the chemokine gradient can become trapped, failing to locate new regions of infected cells in the growing plaque.  The differences in the infection outcomes between the three strains illustrate the effect of the infection spreading faster than the chemokine maxima can move in the case of pH1N1.  The velocity of the chemokine maxima positions depends on a combination of many factors, including the chemokine diffusion, decay, delay, and secretion rates (Table \ref{tab:parameters}).

% Taken together, the delayed response and low proportion of virus secreting cells contribute strongly to the failure of T cells clearing the infection.  Furthermore, increasing chemokine production fails to guide T cells to the actively secreting epithelial cells in the runaway pH1N1 infection (Fig.~\ref{fig:psensitivity}), suggesting that the effect of high viral production rates dominate increased chemokine production rates.

@@

We performed two sensitivity analyses of sixteen model parameters to identify those that have an effect on determining model outcomes (S2.3 - S2.4). We studied all but five of the parameters listed in Tables~\ref{tab:strains} and \ref{tab:parameters}, excluding T cell radius, T cell sensitivity to chemokine, T cell onset time, IgM onset time, and the viral production rate.  T cell radius is known \citep{abbas2011cellular} and does not factor strongly into the spatial model's implementation.  T cell sensitivity was set to mimic a realistic threshold of response to cytokine (see S2.1 for a more complete explanation of how T cell sensitivity was determined).  The two onset parameters are known \citep{Diamond2003}.  Viral production is an independent variable, as discussed above, and determined by each of the three strains \citep{Mitchell2011}.

Each of the remaining sixteen parameters was varied independently (over multiple orders of magnitude when appropriate) for each viral strain using a one-factor-at-a-time (OFAT) approach, and the model was rerun once for each parameter setting (S2.3).  Results were compared for the quantitative size of the peak infection and the number of uncleared infected cells at day 10 p.i.  Based on these runs, we classified each parameter for each strain as follows (Table~\ref{tab:sensitivity}): \textit{stable} parameters are those that do not affect the model behavior over any of the different values tested; \textit{bounded stable} parameters are stable across a biologically plausible range of values but model behavior diverges at some threshold; \textit{peak change} parameters affect the peak size of the infection, but do not change the ultimate number of infected cells at the end of the run; and \textit{sensitive} parameters substantively change the outcome of the simulation within biologically plausible ranges.

Table \ref{tab:sensitivity} shows that four parameters are sensitive: viral response to IgM, infectivity, viral decay rate, and viral diffusion.  In addition, Figure \ref{fig:variance} shows that viral production rate is a sensitive parameter, varying among the three panels.  The four parameters (infectivity, viral decay rate, viral diffusion rate, and viral production rate) are specific viral characteristics.  Further, viral response to IgM, as coded in the model, directly affects viral decay rate and can be considered a viral parameter.  None of the T cell parameters nor the chemokine parameters has a strong effect on the model's prediction of control or lack of control of viral replication.  This result supports the hypothesis that virus dynamics dominate the course of an influenza infection.

Partial rank correlation coefficient (PRCC) analysis was performed on the sH1N1 strain as described in \cite{Marino2008} using the same 16 parameters plus the viral secretion rate to determine the strength and significance of the parameters' partial correlation with the model output (S2.4).  The  five viral parameters are confirmed by the PRCC analysis to be significantly correlated with model output (Table \ref{tab:sensitivity}).  Consistent results between the two approaches suggests the less computationally intensive OFAT analysis is also accurate for aH5N1 and pH1N1.

@@

Our sensitivity analysis suggests that viral parameters are the most sensitive and have the largest effect on the course of the infection (Table~\ref{tab:sensitivity}).  Here we examine the role of T cells by assessing multiple factors that control the overall T cell response.  We seek to identify the maximum amount of time virus-secreting cells may produce virus while still allowing the infection to be cleared.  We name this threshold the `window of control'.  This window can be defined as the combination of the time of T cell arrival at a virus-secreting cell ($T_{arr}$), the time it takes for the T cell to induce apoptosis ($T_{kill}$) and the time for the infected cell to apoptose ($T_{apop}$).  We examine how much virus a virus-secreting cell can produce inside this window.  Values above one suggest a growing infection, while values below one suggest infection control.

@@

$p$, $T_{kill}$, and $T_{apop}$ are model parameters (Table \ref{tab:parameters}).  $E$ incorporates both spatial effects of viral diffusion and temporal effects of viral decay.  As the infected plaque expands, some secreted virus diffuses to cells that are already infected or dead (the existing plaque) and do not contribute to new infections.  This effect is minimal in the beginning when the edge of the plaque is small and tightly curved.  As the plaque expands, its radius grows and the edge approaches the limiting case of a straight line when 50\% of the virus could be expected to diffuse over healthy cells and 50\% over infected or dead cells.  In the presence of healthy cells, virions infect the cells at a rate of $0.5/h$ and decay at a rate of $0.42/h$ in the presence of IgM after day 4 p.i.  (Table \ref{tab:parameters}).  This suggests that 50\% of free virus contacts healthy cells and 54\% of that virus succeeds in infecting a healthy cell, implying that $E \approx 0.27$ (see S2.5 for a more complete explanation of how $E$ was determined).  Solving for $T_{arr}$ where $R_0 < 1$ gives:
